Monte Carlo Corrections for Bayesian Methods

### 2017-05-19 papers/general_method/general.tex
Adding moreeee

So let us modify our imperfect toy model. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$. Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations.
\begin{align}
\mathcal{L} &= \frac{ \mathcal{N}(x|\mu, \sigma) \mathcal{N}(y|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients. One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area.

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous example. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$  we estimate that the average contribution from $\beta y$ is around $4$ (which is close to the correct value of $6$), and from this our analytic correction to our likelihood is
\begin{align}
w = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right]
\end{align}
Further, let us explicitly break our likelihood into two parts, $\mathcal{L} = \mathcal{L}_1 \mathcal{L}_2$, with the parts given by
\begin{align}
\mathcal{L}_1 &= \frac{ \mathcal{N}(x|\mu, \sigma) \mathcal{N}(y|\mu_y, \sigma_y)}{w} \\
\mathcal{L}_2 &= \frac{ w }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
$\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either - each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable. Evaluating $\mathcal{L}_2$ using Monte Carlo integration of $n$ samples, we have
\begin{align}
\mathcal{L}_2 = \frac{ w n }{\sum_{i=1}^{n} \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X_i|\mu, \sigma) \mathcal{N}(Y_i|\mu_y, \sigma_y)}
\end{align}

>>>>

So let us modify our imperfect toy model. Instead of observing just one variable, $x$, we also observe a new independent variable, $y$, which is drawn from its own distribution $y \sim \mathcal{N}(\mu_y, \sigma_y)$. Our selection efficiency can now become a combination of $x$ and $y$, such that we only observe events that satisfy $x + \beta y > \alpha$, giving $P(S|x,y,\theta) = \mathcal{H}(x + \beta y - \alpha)$.  Our likelihood for such a toy model becomes now the combination of probabilities for observing both $x$ and $y$, with the denominator becoming an integral over all possible $X$ and $Y$ observations subject to our selection effects.
\begin{align}
\mathcal{L} &= \frac{ \mathcal{N}(x|\mu, \sigma) \mathcal{N}(y|\mu_y, \sigma_y)}
{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}
\end{align}
Assume that we cannot solve this integral analytically, and must resort to numeric solutions. These often clash with sampling methods, especially for high dimensional integrals. Inserting Monte Carlo integration into fitting algorithms can drastically slow them down, and algorithms such as Hamiltonian MCMC that require continuous surfaces can easily fail on surfaces that fluctuate from Monte Carlo integration. Even by fixing the samples used in MC integration (thereby giving a continuous surface), the complexity of the surface derivatives will pose almost insurmountable problems for any algorithms that utilise surface gradients. One solution is to find an approximate, analytic correction we can utilise in our fitting algorithm which seeks to shift the region of parameter space sampled by the sampler closer to the correct area.

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous correction from Section \ref{sec:imperfect}. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$ (perhaps we believe it is approximately $20$) we estimate that the average contribution from $\beta y$ is around $20\beta = 4$ (which is close to the correct value of $6$), and from this our analytic correction to our likelihood is
\begin{align}
w = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right].
\end{align}
Further, let us explicitly break our likelihood into two parts, $\mathcal{L} = \mathcal{L}_1 \mathcal{L}_2$, with the parts given by
\begin{align}
\mathcal{L}_1 &= \frac{ \mathcal{N}(x|\mu, \sigma) \mathcal{N}(y|\mu_y, \sigma_y)}{w} \\
\mathcal{L}_2 &= \frac{ w }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
$\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either - each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable. Evaluating $\mathcal{L}_2$ using Monte Carlo integration of $n$ samples, we have
\begin{align}
\mathcal{L}_2 = \frac{ w n }{\sum_{i=1}^{n} \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X_i|\mu, \sigma) \mathcal{N}(Y_i|\mu_y, \sigma_y)}
\end{align}
